Mastery Learning in Practice:\\ A Descriptive Analysis of Data from the Cognitive Tutor Algebra I Effectiveness Trial

### 2017-10-19 paper.Rnw
more modeling...

xtable(tab,caption='Median numbers of hours, problems, sections, and units worked by each student in the dataset in the two years of the study. Students with no usage data were excluded',label='tab:medUsage',digits=c(1,2,0,1,0))
@

@@

print(unitsStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
@
\caption{Boxplots of the number of units of Cognitive Tutor
  software each student worked, by year and state. Students
  working more than 55 units (\Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)}) of \Sexpr{nrow(secByStud)}) and students
  with no usage data (\Sexpr{sum(is.na(secByStud[['numUnit']]))})
  were excluded.}
\label{fig:unitsByStud}
\end{figure}

@@

\subsection{How Do Reassignment Patterns Vary?}
Teachers alone control reassignment.
Nevertheless, the factors influencing student reassignment vary at a
number of levels.
For instance, state and district standards may prod teachers into
reassigning students to particular units.
Some prinicpals may encourage teachers to adhere to the official
curriculum and avoid reassignment.
Some students may be more prone to reassignment than others.
Certain units in the CTAI curriculum may be harder than others,
causing students to tarry and teachers to reassign.
Finally, a host of other factors, at these levels and others, may spur reassignment.

@@

@
\caption{Results from a set of eight multilevel logistic regressions
  predicting section reassignment. For each year, in the entire sample (``Overall'')
  and in the three states with the largest numbers of reassignments (Texas,
  Kentucky, and Michigan), we regressed a binary variable indicating
  whether a section ended in reassignment on random intercepts for
  school, class, student, and unit, and in the overall case, for state
  as well, and recorded their variance. The residual variance was set
  as the variance of the standard logistic distribution,
  $\pi/3$. These bar charts give the proportion of the total variance
  attributable to each random effect.}
\label{fig:vc}
\end{figure}

@@

To address these questions, we focus on the units in the standard
Algebra I sequence.
Figures \ref{fig:trans1} and \ref{fig:trans2} give transition plots
for reassignment in the two study years.
On the left of each figure, marked ``From,'' are the top ``sending''
units: the units that students were reassigned \emph{from} at least 30
times.
They are listed in the order they appear in the standard Algebra I curriculum.
On the right, under ``To,'' are the top ``receiving'' units: the units
that students were reassigned \emph{to} at least 30 times.
For completeness, each of the top two receivers for each sending unit,
and each of the top two senders for each receiving unit were also included.
Finally, all of the sending units were also listed on the left, in the
receiving column.

The arrows in the plot represent reassignments.
An arrow from a sending unit to a receiving unit indicates
reassignment from the former to the latter.
The thickness of the arrows represents the proportion of reassignments
originating in the sending unit whose destination was the receiving unit.
The darkness of the arrows represents the number of such reassignments.
For instance, in Figure \ref{fig:trans1}, the thick line from
``4−Quadrant Linear Graphs'' on the left to ``4−Quadrant Linear
Graphs'' on the right is fairly thick, since

@@

\begin{figure}
  \begin{subfigure}{.7\textwidth}
<<transition1,fig.height=3,fig.width=4.5,dependson='secOrder'>>=

@@

propTotCP <- sum(trans1)/sum(secOrderCP$year==1)
#trans1itionPlot(trans1)

nsec <- nrow(trans1)
uuu <- levels(secOrderCP$unitName)
uuu <- uuu[uuu%in%c(colnames(trans1),rownames(trans1))]
nsec <- length(uuu)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans1)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
for(i in 1:nsec) if(uuu[i]%in%colnames(trans1)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))

@@

@
\end{subfigure}
\begin{subfigure}{.25\textwidth}
<<trans1legend,fig.width=1.6,fig.height=3>>=
par(mar=c(.1,.1,.1,.1))
plot(1,1,xlim=c(0,2),ylim=c(0,11),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
text(1,10,expression(underline(bold(paste(Legend)))))
text(1,8,'Thickness: Percent')
text(.5,7,'25%')
text(.5,6,'50%')
text(.5,5,'75%')
arrows(c(1.5,1.5,1.5),c(7,6,5),c(2,2,2),lwd=5*c(.25,.5,.75))
text(1,4,'Darkness: Number')
text(.5,3,'10')
text(.5,2,'25')
text(.5,1,'50+')
arrows(c(1.5,1.5,1.5),c(3,2,1),c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))

@
\caption{Reassignment transition plot for study year 1. The units on
  the left of the plot are the Algebra I units that ended in reassignment at
  least 30 times in year 1. The units on the right are those that were
  the destination of at least 30 reassignments. Also included on the
  left are the top two sending units for each of the units on the
  right, and also included on the right are all the units on the left,
  along with their top two receiving units. The units are listed
  following their order in the standard Algebra I curriculum. There is an arrow from a
  unit on the left to a unit on the right if a student was assigned
  from the unit on the left to the one on the right. The thickness of
  the arrows is proportional to the percentage of all reassignments from
  the sending unit that ended in that receiving unit. The darkness of
  the arrows is proportional to the number of reassignments from the
  sending unit to the receiving unit.}
\label{fig:trans1}
\end{figure}

\begin{figure}
<<transition2,fig.height=5,fig.width=6.5,dependson='secOrder'>>=

@@

propTotCP <- sum(trans2)/sum(secOrderCP$year==1)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans2)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
for(i in 1:nsec) if(uuu[i]%in%colnames(trans2)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))

for(rr in rownames(trans2))
 for(cc in colnames(trans2)){
#     if(trans2[rr,cc]>20){
      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans2[rr,cc]/sum(trans2[rr,]),
       col=rgb(0,0,0,alpha=ifelse(trans2[rr,cc]>50,1,trans2[rr,cc]/50)))
}
@
\caption{Reassignment transition plot for study year 2.}
\label{fig:trans2}
\end{figure}

\subsection{Does Reassignment Depend on Student Ability?}

<<xirtMods,include=FALSE,dependson=c('data','setup')>>=
## don't worry about nesting, etc--we aren't reporting standard errors anyway
cpDat <- data%>%filter(!is.na(status))%>%group_by(field_id,unit,section,state,year,classid2,schoolid2)%>%
    summarize(cp=max(status)=='changed placement')

### get imputed xirt
##
h1 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H1_algebra_rcal_20121119_fieldid.csv')
h2 <- read.csv('~/Box Sync/CT/data/RANDstudyData/H2_algebra_rcal_20121119_fieldid.csv')

h2 <- h2[!h2$field_id%in%h1$field_id,]

stud <- rbind(h1[,c(174:194)],h2[,c(176:196)])
stud <- stud[stud$field_id%in%cpDat$field_id,]
stud <- stud[order(stud$field_id),]
stud$xirt <- rowMeans(stud[,grep('Exirt2_',names(stud))],na.rm=TRUE)

cpDat$xirt <- stud$xirt[match(cpDat$field_id,stud$field_id)]

cpModSimp <- glm(cp~splines::ns(xirt,5)*year,data=cpDat)

cpModSt <- list()
for(st in c('TX','KY','MI','LA','CT')){
    cpModSt[[st]] <- update(cpModSimp,data=subset(cpDat,state==st))
}

\begin{figure}
<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup','xirtMods')>>=
#load('cpXirtMods.RData')

xirtDat <- cpDat%>%group_by(field_id,state,year,classid2,schoolid2,xirt)%>%
    summarize(nsec=n_distinct(section,unit),propCP=mean(cp,na.rm=TRUE))

for(st in names(cpModSt)) xirtDat$fitted[xirtDat$state==st] <-
                                  predict(cpModSt[[st]],xirtDat[xirtDat$state==st,],
                                         type='response')

xirtDat <- addOverallState(xirtDat)

xirtDat$fitted[xirtDat$state=='Overall'] <-
    predict(cpModSimp,xirtDat[xirtDat$state=='Overall',],
            type='response')

ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red',size=1)+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.5))

@
\end{figure}

\begin{figure}
<<cp-xirtClassroom1,dependson='xirtCP',fig.height=7,fig.width=6.5>>=
bbb2 <- data%>%filter(status=='changed placement')%>%group_by(classid2,year,state)%>%summarize(ncp=n_distinct(field_id,section,unit),nstud=n_distinct(field_id),nsec=n_distinct(section))

cpcls1 <- xirtDat%>%
    filter(classid2%in%bbb2$classid2[bbb2$year==1 & bbb2$ncp>=20] & state!='Overall')%>%
    arrange(state)
cpcls1$classid2 <- factor(cpcls1$classid2,levels=unique(cpcls1$classid2))

cpcls1 <- cpcls1%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()

ggplot(cpcls1,aes(xirt,propCP,size=nsec,color=state))+
    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
    ylim(0,1)

@
\end{figure}

\begin{figure}
<<cp-xirtClassroom2,dependson='cp-xirtClassroom1',fig.height=7,fig.width=6.5>>=

cpcls2 <- xirtDat%>%
    filter(classid2%in%bbb2$classid2[bbb2$year==2 & bbb2$ncp>=20] & state!='Overall')%>%
    arrange(state)
cpcls2$classid2 <- factor(cpcls2$classid2,levels=unique(cpcls2$classid2))

cpcls2 <- cpcls2%>%group_by(classid2)%>%mutate(maxCP=max(propCP,na.rm=TRUE),propCP=propCP/maxCP)%>%ungroup()

ggplot(cpcls2,aes(xirt,propCP,size=nsec,color=state))+
    geom_point()+geom_smooth(size=1,method='lm')+facet_wrap('classid2')+scale_size_continuous(range=c(0.2,1))+
    ylim(0,1)

@
\end{figure}

@@

\end{document}

>>>>

xtable(tab,caption='Median numbers of hours, problems, sections, and units worked by each student in the dataset in the two years of the study. Students with no usage data were excluded.',label='tab:medUsage',digits=c(1,2,0,1,0))
@

@@

print(unitsStateYear <- ggplot(filter(secByStud2,state!='NJ'),aes(Yr,numUnit))+geom_boxplot(outlier.shape=NA)+geom_jitter(data = function(x) dplyr::filter_(x, ~ outlier), width=0.2)+facet_grid(~state)+labs(x='',y='# Units Worked Per Student')+coord_cartesian(ylim=c(0,55)))
@
\caption{Boxplots of the number of units of Cognitive Tutor
  software each student worked, by year and state. Students
  working more than 55 units (
  \Sexpr{sum(secByStud[['numUnit']]>55,na.rm=TRUE)} of
  \Sexpr{nrow(secByStud)}) and students
  with no usage data (\Sexpr{sum(is.na(secByStud[['numUnit']]))})
  were excluded.}
\label{fig:unitsByStud}
\end{figure}

@@

\subsection{How Do Reassignment Patterns Vary?}

@@

@
\caption{Results from a set of eight multilevel logistic regressions
  predicting section reassignment. For each year, in the entire sample (``Overall'')
  and in the three states with the largest numbers of reassignments (Texas,
  Kentucky, and Michigan), we regressed a binary variable indicating
  whether a section ended in reassignment on random intercepts for
  school, class, student, and unit, and in the overall case, for state
  as well, and recorded their variance. The residual variance was set
  as the variance of the standard logistic distribution,
  $\pi/3$. These bar charts give the proportion of the total variance
  attributable to each random effect.}
\label{fig:vc}
\end{figure}
Teachers alone control reassignment.
Nevertheless, the factors influencing student reassignment vary at a
number of levels.
For instance, state and district standards may prod teachers into
reassigning students to particular units.
Some prinicpals may encourage teachers to adhere to the official
curriculum and avoid reassignment.
Some students may be more prone to reassignment than others.
Certain units in the CTAI curriculum may be harder than others,
causing students to tarry and teachers to reassign.
Finally, a host of other factors, at these levels and others, may spur reassignment.

@@

To address these questions, we focus on the units in the standard
Algebra I sequence.
Figures \ref{fig:trans1} and \ref{fig:trans2} give transition plots
for reassignment in the two study years.
On the left of each figure, marked ``From,'' are the top ``sending''
units: the units that students were reassigned \emph{from} at least 30
times.
They are listed in the order they appear in the standard Algebra I curriculum.
On the right, under ``To,'' are the top ``receiving'' units: the units
that students were reassigned \emph{to} at least 30 times.
For completeness, each of the top two receivers for each sending unit,
and each of the top two senders for each receiving unit were also included.
Finally, all of the sending units were also listed on the left, in the
receiving column.
All in all, \Sexpr{round(propTotCP1*100)}\% of the first-year reassignments and
\Sexpr{round(propTotCP1*100)}\% of the second-year reassignments are
captured in the figures.

The arrows in the plot represent reassignments.
An arrow from a sending unit to a receiving unit indicates
reassignment from the former to the latter.
The thickness of the arrows represents the proportion of reassignments
originating in the sending unit whose destination was the receiving unit.
The darkness of the arrows represents the number of such reassignments.
For instance, in Figure \ref{fig:trans1}, the arrow from
``4−Quadrant Linear Graphs'' on the left to ``4−Quadrant Linear
Graphs'' on the right is fairly thick, since
\Sexpr{round(trans1['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']/
  sum(trans1['4-Quadrant Linear Graphs',])*100)}\%
of the reassignments begun in ``4-Quadrante Linear Graphs'' end in the
same unit.
Yet it is also fairly faint, since it only represents
\Sexpr{trans1['4-Quadrant Linear Graphs','4-Quadrant Linear Graphs']}
reassignments.

@@

\begin{figure}
  \begin{subfigure}{5in}
<<transition1,fig.height=3,fig.width=5,dependson='secOrder'>>=

@@

propTotCP1 <- sum(trans1)/sum(secOrderCP$year==1)

uuu <- levels(secOrderCP$unitName)
uuu <- uuu[uuu%in%c(colnames(trans1),rownames(trans1))]
nsec <- length(uuu)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans1)) text(1,nsec-i+1,uuu[i],adj=c(1,NA),cex=0.8)
for(i in 1:nsec) if(uuu[i]%in%colnames(trans1)) text(2,nsec-i+1,uuu[i],adj=c(0,NA),cex=0.8)

@@

@
\end{subfigure}
\begin{subfigure}{1in}
<<trans1legend,fig.width=1,fig.height=3,dependson='transition1'>>=
par(mar=c(.1,.1,.1,.1))
plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
conv <- (nsec+2)/11
text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
text(1,9*conv,'Thickness:\n Percent',cex=0.8)
text(.5,7.5*conv,'25%')
text(.5,6.5*conv,'50%')
text(.5,5.5*conv,'75%')
arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
text(1,4.2*conv,'Darkness:\n Number',cex=0.8)
text(.5,3*conv,'10')
text(.5,2*conv,'25')
text(.5,1*conv,'50+')
arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))

@
\end{subfigure}
\caption{Reassignment transition plot for study year 1. The units on
  the left of the plot are the Algebra I units that ended in reassignment at
  least 30 times in year 1. The units on the right are those that were
  the destination of at least 30 reassignments. Also included on the
  left are the top two sending units for each of the units on the
  right, and also included on the right are all the units on the left,
  along with their top two receiving units. The units are listed
  following their order in the standard Algebra I curriculum. There is an arrow from a
  unit on the left to a unit on the right if a student was assigned
  from the unit on the left to the one on the right. The thickness of
  the arrows is proportional to the percentage of all reassignments from
  the sending unit that ended in that receiving unit. The darkness of
  the arrows is proportional to the number of reassignments from the
  sending unit to the receiving unit.}
\label{fig:trans1}
\end{figure}

Inspecting the figures shows that the most common pattern is for
students to be reassigned to the next unit in the curriculum.
On the other hand, it is rare for students to be reassigned

\begin{figure}
  \begin{subfigure}{5in}
<<transition2,fig.height=5,fig.width=5,dependson='secOrder'>>=

@@

propTotCP2 <- sum(trans2)/sum(secOrderCP$year==1)

@@

for(i in 1:nsec) if(uuu[i]%in%rownames(trans2)) text(1,nsec-i+1,uuu[i],adj=c(1,NA),cex=0.8)
for(i in 1:nsec) if(uuu[i]%in%colnames(trans2)) text(2,nsec-i+1,uuu[i],adj=c(0,NA),cex=0.8)

for(rr in rownames(trans2))
 for(cc in colnames(trans2)){
#     if(trans2[rr,cc]>20){
      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans2[rr,cc]/sum(trans2[rr,]),
       col=rgb(0,0,0,alpha=ifelse(trans2[rr,cc]>50,1,trans2[rr,cc]/50)))
}
@
\end{subfigure}
\begin{subfigure}{1in}
<<trans2legend,fig.width=1,fig.height=3,dependson='transition2'>>=
par(mar=c(.1,.1,.1,.1))
plot(1,1,xlim=c(0,2),ylim=c(0,nsec+2),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
conv <- (nsec+2)/11
text(1,nsec+1.5,expression(underline(bold(paste(Legend)))))
text(1,9*conv,'Thickness:\n Percent',cex=0.8)
text(.5,7.5*conv,'25%')
text(.5,6.5*conv,'50%')
text(.5,5.5*conv,'75%')
arrows(c(1,1,1),c(7.5,6.5,5.5)*conv,c(2,2,2),lwd=5*c(.25,.5,.75))
text(1,4.2*conv,'Darkness:\n Number',cex=0.8)
text(.5,3*conv,'10')
text(.5,2*conv,'25')
text(.5,1*conv,'50+')
arrows(c(1,1,1),c(3,2,1)*conv,c(2,2,2),col=rgb(0,0,0,alpha=c(1/5,.5,1)))

@
\end{subfigure}
\caption{Reassignment transition plot for study year 2.}
\label{fig:trans2}
\end{figure}

\section{Effects of Reassignment}
The goal of CTAI is to help students learn Algebra, so
the most important questions about reassignment are about its effect on learning.
Although the data from this study came from a randomized trial, it was
CTAI as a whole that was randomized, not individual behaviors within CTAI.
Specifically, student reassignment was not randomized.
Therefore, claims regarding the causal effects of reassignment on
learning require strong untestable assumptions that are unlikely to be
true.
As in all observational studies, this includes the assumption that all
counfounding variables---variables that predict both reassignment and
learning---have been measured well and modeled correctly.
Further complicating matters, although reassignment itself is a
well-defined process, in practice it can take many forms, as we have
endevoured to show.
There is no reason to expect the effects of reassignment to be
the same regardless of wheather the teacher used it to help lagging
students catch up, to allocate time to important topics, or for some
other reason.

All that said, observational estimates of reassignment's average
effects can be valuable, if interpreted cautiously.
They may help guide research

\subsection{Effects on Subsequent Mastery}
<<>>=
@

\subsection{Effects on Post-Test Scores}

@@

\end{document}
<<eval=FALSE>>=

 nextDat <- secOrder%>%filter(!is.na(status))%>%group_by(field_id,classid2,year,state)%>%
  summarize(naf=sum(is.na(prevStatus) & status=='final_or_incomplete')/sum(is.na(prevStatus)),
   nacp=sum(is.na(prevStatus) & status=='changed placement')/sum(is.na(prevStatus)),
   naprom=sum(is.na(prevStatus) & status=='promoted')/sum(is.na(prevStatus)),
   nagrad=sum(is.na(prevStatus) & status=='graduated')/sum(is.na(prevStatus)),
   ff=sum(prevStatus==1 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
   fcp=sum(prevStatus==1 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
   fprom=sum(prevStatus==1 & status=='promoted',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
   fgrad=sum(prevStatus==1 & status=='graduated',na.rm=TRUE)/sum(prevStatus==1,na.rm=TRUE),
   cpf=sum(prevStatus==2 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
   cpcp=sum(prevStatus==2 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
   cpprom=sum(prevStatus==2 & status=='promoted',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
   cpgrad=sum(prevStatus==2 & status=='graduated',na.rm=TRUE)/sum(prevStatus==2,na.rm=TRUE),
   promf=sum(prevStatus==3 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
   promcp=sum(prevStatus==3 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
   promprom=sum(prevStatus==3 & status=='promoted',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
   promgrad=sum(prevStatus==3 & status=='graduated',na.rm=TRUE)/sum(prevStatus==3,na.rm=TRUE),
   gradf=sum(prevStatus==4 & status=='final_or_incomplete',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
   gradcp=sum(prevStatus==4 & status=='changed placement',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
   gradprom=sum(prevStatus==4 & status=='promoted',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
   gradgrad=sum(prevStatus==4 & status=='graduated',na.rm=TRUE)/sum(prevStatus==4,na.rm=TRUE),
   npna=sum(is.na(prevStatus)),npf=sum(prevStatus==1,na.rm=TRUE),npcp=sum(prevStatus==2,na.rm=TRUE),
    npprom=sum(prevStatus==3,na.rm=TRUE),npgrad=sum(prevStatus==4,na.rm=TRUE))
